% -*- outline-regexp: "%--* ";  -*-
\documentclass[a4]{article}
%- Packages
%% Unicode...
% \usepackage{ucs}
% \usepackage[utf8x]{inputenc}
% \usepackage{textcomp}

\usepackage{url,lineno}

%-- Mathematics
\usepackage{mathtools}
% \usepackage{accents}
\usepackage{amsmath}
% \usepackage{amsthm}
% \usepackage{amssymb}
% \usepackage{mathtools}
% \usepackage{enumitem}
\usepackage{natbib}
\usepackage{lmodern}

%-- Algorithms
\usepackage{algorithm}
% \usepackage[defblank]{paralist}
%% \usepackage{algorithmicx}
\usepackage{algpseudocode}

%-- Lists
\usepackage[ampersand]{easylist}

%-- Fonts
\usepackage[T1]{fontenc}
\usepackage{calligra}
\usepackage{chancery}


%- Flags, such as \linenumbers
%%% Uncomment to use
% \linenumbers


%- Fonts
\def\origrm{rmdefault}
\renewcommand{\rmdefault}{origrm}

\newcommand{\resetfonts}[0]{\fontencoding{\encodingdefault}\fontfamily{\familydefault}\fontseries{\seriesdefault}\fontshape{\shapedefault}\selectfont}
\def\calli#1{{\calligra #1}}
\def\chanc#1{{\renewcommand{\rmdefault}{pzc}\rm #1 \renewcommand{\rmdefault}{origrm}}}

%- Maths alphabets
\DeclareMathAlphabet{\mathsfsl}{OT1}{cmss}{m}{sl}
\DeclareMathAlphabet{\mathpzc}{OT1}{pzc}{m}{it}

%- Maths operators

%- Load extra configuration before the newcommands 
% \message{Loading Manifest ==========================================}
% \input{manifest.tex}
% \message{Finished Manifest =========================================}

%- Environments

\newenvironment{indented}{\begin{adjustwidth}{24pt}{}}{\end{adjustwidth}}

%% declares the character degreesC (U+2103) to map to the \textcelsius function
%% \DeclareUnicodeCharacter{"2103}{\textcelsius}

% Newcommands
\newcommand{\mt}[1]{\mathtt{#1}}
\newcommand{\mB}[1]{\mathbb{#1}}
\newcommand{\mb}[1]{\mathbf{#1}}
\newcommand{\mf}[1]{\mathfrak{#1}}
\newcommand{\mc}[1]{\mathpzc{#1}}
% \newcommand{\mc}[1]{\mathcal{#1}}

\newcommand{\filename}[1]{\texttt{#1}}
\newcommand{\statevariable}[1]{\textit{#1}}
\newcommand{\rclass}[1]{\textsf{<#1>}}
\newcommand{\method}[1]{\textbf{\textit{#1}}}
\newcommand{\Function}[1]{\textbf{\slfamily{#1}}}

\newcommand{\landscape}[0]{\rclass{landscape}}
\newcommand{\environment}[0]{\rclass{environment}}
\newcommand{\ecoservice}[0]{\rclass{ecoservice}}
\newcommand{\patch}[0]{\rclass{patch}}
\newcommand{\patches}[0]{{\patch}es}
\newcommand{\dynamicpatch}[0]{\rclass{dynamic-patch}}
\newcommand{\dynamicpatches}[0]{{\dynamicpatch}es}
\newcommand{\diffeqsystem}[0]{\rclass{diffeq-system}}
\newcommand{\populationsystem}[0]{\rclass{population-system}}
\newcommand{\polygonshape}[0]{\rclass{polygonshape}}
\newcommand{\circleshape}[0]{\rclass{circleshape}}

\newcommand{\extset}[0]{\method{extset!}
\newcommand{\extget}[0]{\method{extget}

\newcommand{\make}[0]{\Function{make}}
\newcommand{\makeagent}[0]{\Function{make-agent}}
\newcommand{\makeobject}[0]{\Function{make-object}}


\begin{document}
\title[Remodel]{}

% Some useful definitions...

\section{Overview}

\section{Extended \SCLOS}

\SCLOS is an object system closely modelled on the Common Lisp Object
System (\CLOS); \Remodel uses \SCLOS with extensions: agents are a
special kind of object, and require a significant amount of
infrastructure that objects do not need, such as the ability to run
the agent's ``model body'' which simulates its activity through a
timestep. Where \SCLOS provides the \make function, \Remodel also
provides \makeagent and \makeobject. The extensions provide features
which a simple \SCLOS object does not have, such as unique object
identifiers, infrastructure which allows agents to run, communicate
with each other and with the kernel, and to migrate to different
representations. \Remodel will raise an error if code tries to make an
agent using \makeobject or to make an object using \makeagent.  It
should be noted that calls to the bare \Function{make} are \emph{not}
trapped or tested, and that \Function{make} should really only be used
in constructor functions which have been written with these issues in
mind. 



    


\section{Discussion of the model structure}

\subsection{Landscape classes}
The classes which represent the ``landscape'' of the model are
ultimately related to the \environment class described in
\filename{remode-classes.scm}. These classes would generally be
responsible for representing physical componenets, such as the
terrain, groundwater and surface water, but they may also be used to
represent elements that might be referred to with mass-nouns, such as
grass, forest or foliage.  

This class maintains the physical boundaries of the environment
(minimum and maximum values for spatial ordinates, a default spot
height, an inner radius and an outer radius which can be used to speed
``presence'' checks, and a ``representation'' which isi usually
something like a polygon, a DEM or some similar object.

The \landscape code consists mainly of a few classes.  Strictly
speaking, a \landscape can include the terrain, and various
\textit{attributes} which are characterised by numbers which can
increase or decrease. The \textit{attributes} are usually encapsulated
in \patches which are spatially anchored regions which (properly)
should not overlap significantly.  There is ready support for
rectangular and hexagonal regions that cover the the domain of the
\landscape in a tesselation, though there are also circular
representations which may be appropriate in simulations where complete
coverage is not required -- these patch shapes obviously don't
tesselate!  A terrain function is also supplied when creating a
landscape. This function typically expects a locus and returns an
altitude relative to the datum  the model.

The attributes mentioned above are implemented as members of the
\ecoservice class.  This is a class which basically maintains some
``value'' associated with the patch, such as the biomass of grass, or
the number of pademelons. It is able to act either as an agent
maintaining the state of some environmental component or as a proxy
for a set of other agents which should be treated as a group (perhaps
a population of rabbits) using \extset and \extget calls. The
water-table in a patch may be modelled using an ecoservice: its
internal recharge model will cause the amount of available water to
increase as a function of its natural recharge rate, while flora in
the patch and local industry which extracts water will reduce the
available water with appropriate calls to secure the water they need.
Ecoservices keep track of the patches for which they are relevant and
\textit{vis-a-versa}. Externally modelled components that are treated
as ecoservices would be polled at the beginning of the ecoservice's
slice of time in order to set the numeric \statevariable{value} slot
in the \ecoservice agent, and at the end of the update step the
corresponding values for the agents polled would be adjusted
appropriately. 

\patches maintain a list of their ecoservices, their spatial location
and footprint, a variable (\textit{notepad} containing a list which
can be used for any cross-tick storage, a list of the patch's
neighbours, and an optional ``caretaker'' function which is called at
each timestep. Neighbours in the list are represented by lists of the
form '\textit{(neighbour-patchagent number | symbol \#!optional
  extras)}'; the \textit{extras} component is not required and
currently not used in any of the submodels.  The numeric second
argument represents the ``length'' of the boundary between the agent
and its neighbour (which can be used to mediate the exchange between
them).

The footprints of the patches are represented either with \circle or
\polygon objects, and the ecoservices are treated as being
equally accessible over the whole region covered by the
footprint, whether they are implemented withing the ecoservices or are
external agents accessed through the \ecoservice agent.

\dynamicpatches are derived from both \patch and \diffeqsystem and
allows the patch to behave as an ensemble which is governed by a
system of differential equations. Interactions between neigbouring
sources and sinks for the ecoservices represented in the system are
supported and worked into the dynamics of a timestep.

There are also simple geometric classes that correspond to ``regions''
such as \polygonshape and \circleshape; while squares and hexagons
feature as tiles in a landscape, they are simply instances of
\polygonshape.

\subsubsection{Constructing landscapes}

The declaration
\begin{verbatim}
  (define HeronIs 
     (make <landscape>
        'name "Heron Island" 
        'default-value (* 0.5 meters)
        'minv (0 0 -5) 'maxv '(0 1000 122)
        'terrain-function heronDEM
        'patch-list (list southwest-heron southeast-heron
                northwest-heron north-heron northeast-heron)
     ))
\end{verbatim}
could also be expressed using the function \Function{make-landscape}
like so:
\begin{verbatim}
   (define HeronIs (make-landscape "Heron Island" 0.5 domain heronDEM 
      (list southwest-heron southeast-heron northwest-heron north-heron northeast-heron)
      )
\end{verbatim}

Both forms are provided because the first is more useful when the
construction of the object is generated from a general configuration
rather than hard coded.  There are similar constructors
for some of the other major classes.







\end{document}
