% prelude.tex (specification of which features in `mathphdthesis.sty' you
% are using, your personal information, and your title & abstract)

% Specify features of `mathphdthesis.sty' you want to use:
\titlepgtrue%  main title page (required)
\signaturepagetrue%  page for declaration of originality (required)
%\copyrighttrue%  copyright page (required)
\altcopyrighttrue%  copyright page (required)
\abswithesistrue%  abstract to be bound with thesis (optional)
\acktrue%  acknowledgments page (optional)
\tablecontentstrue%  table of contents page (required)
\tablespagetrue%  table of contents page for tables (required only if you have tables)
\figurespagetrue%  table of contents page for figures (required only if you have figures)

\title{ADAPTIVE SIMULATION MODEL CONFIGURATION}%  use all capital letters
\author{Randall Gray}%  use mixed upper & lower case
\prevdegrees{B.A. (Flinders University of South Australia)}%  Used to specify your previous degrees...use mixed upper & lower case
\advisor{Dr Simon Wotherspoon}%  example: Professor Lawrence K. Forbes
\dept{Mathematics}%  your academic department
\submitdate{October, 2016}%  month & year of your thesis submission

\input{manifest}

\newcommand{\abstextwithesis}
{Models of complex systems may be improved by allowing the way they
represent component parts to change as the state of the model and its
components move through their state-spaces.

An argument is made that both process fidelity and computational
efficiency can be improved. This is demonstrated by a simple adaptive
model that changes its basic representation for a population
encountering contaminants performs better than either of the
conventional forms alone.  In this example, there was a substantial
decrease in runtime relative to a purely \istateC\ model, and the
adaptive model performed as well as the \istateC\ model, while
the purely \pstate\ implementation suffered from errors arising from
the ``blurring'' of contact with contaminants through the distributed
population.  This example develops the notion of model representations
maintaining \istateC\ data of other representations so that
representations can be recovered with minimal error.

Deciding when to change representations of components is addressed
initially in a paper exploring a possible set of dynamics associated
with a simple, hypothetical model of a seven component ecosystem.  The
components of the system are described as \istateC\ models or as \pstate\
models, and a mechanism for determining when to change representations
is outlined.  

To support the analysis and selection of representations, a
metric-space with the properties of a commutative ring is defined. The
elements of the metric-space are trees which can be used to encode the
structural character of a set of submodels which comprise the model of
a system, and to allow a targetted transition from one configuration
to another.

Finally, a complete model is presented.  The model closely follows the
hypothetical example and was designed as a reference model, and is
freely available
on \texttt{https://github.com/snarkypenguin/Mutans.git}
This implementation demonstrates the proposed dynamic configuration
management, the maintenance of the state of superceded submodel
representations, and the support structures needed to implement models
of this type.
}




\newcommand{\acknowledgement}
{My deepest thanks go to my very patient wife, Anne, who coped with
far too many years with me firmly attached to my keyboard.  Thanks
also go my supervisor, Simon, who was never short of optimism and
encouragement. I would also like to acknowlege my debt and gratitude
to the late Dr William Cornish of Flinders University, who kindled my
love of mathematics in 1985.}

% Take care of things in `mathphdthesis.sty' behind the scenes.
% Basically just does a check of all the fields that have been activated
% above and fills out the appropriate pages and adds them to the thesis.
%-\beforepreface
%-\afterpreface
